Different parameterizations for one factor ANOVA

### 2022-07-07 text/22/different-parameterizations.Rmd
Finished different-parameterizations.Rmd

```{r}
means <- c(50, 52, 60, 70)
mean(means)
y <- 
  rep(means, each=3) +
  rep(c(0, 1, -1), 4)
g <- rep(1:4, each=3)
df <- cbind(g, y)
```

@@

```{r}
x <- cbind(x1, x2, x3, x4) 
x
```

@@

```{r}
lm(y~x1+x2+x3+x4-1)
```

@@

$\hat{\beta}=(X'X)^{-1}X'$

@@

```{r}
inv(t(x)%*%x)%*%t(x) %>% fractions
```

@@

```{r}
x <- cbind(x0, x1, x2, x3) 
x
```

@@

```{r}
lm(y~x0+x1+x2+x3-1)
```

```{r}
inv(t(x)%*%x)%*%t(x) %>% fractions
```

@@

@@

So the slope associated with X1 requires you to hold X2 and X3 constant. You can hold them constant at 0 or hold them constant at 1, but the former is simpler to understand. THink of it as moving from one of the first three rows of your X matrix.

```{r}
x[1:3, ]
```

@@

```{r}
x[4:6, ]
```

@@

>>>>

```{r}
means <- c(52, 54, 62, 72)
mean(means)
y <- 
  rep(means, each=3) +
  rep(c(0, 1, -1), 4)
g <- rep(1:4, each=3)
df <- cbind(g, y)
```

@@

```{r}
A <- cbind(x1, x2, x3, x4) 
A
```

@@

```{r}
beta_A <- coef(lm(y~x1+x2+x3+x4-1))
beta_A
```

@@

$(X'X)^{-1}X'$

@@

```{r}
inv(t(A)%*%A)%*%t(A) %>% fractions
```

@@

```{r}
B <- cbind(x0, x1, x2, x3) 
B
```

@@

```{r}
beta_B <- coef(lm(y~x0+x1+x2+x3-1))
beta_B
```

```{r}
inv(t(B)%*%B)%*%t(B) %>% fractions
```

@@

In other words,

$\hat{\beta}_0=\bar{Y}_1$

$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_1$

$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_1$

$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_1$

@@

So the slope associated with X1 requires you to hold X2 and X3 constant. You can hold them constant at 0 or hold them constant at 1, but the former is simpler to understand. Think of it as moving from one of the first three rows of your X matrix.

```{r}
B[1:3, ]
```

@@

```{r}
B[4:6, ]
```

@@

### Deviation coding

If you tried to code the difference in means directly, you get a third result.A common way to parameterize the ANOVA model is to use an intercept term and indicator variables for all but one of the groups.

```{r}
x0 <- rep(1, 12)
x1 <- as.numeric(g==2) - as.numeric(g==1)
x2 <- as.numeric(g==3) - as.numeric(g==1)
x3 <- as.numeric(g==4) - as.numeric(g==1)
```

```{r}
D <- cbind(x0, x1, x2, x3) 
D
```

```{r}
beta_D <- coef(lm(y~x0+x1+x2+x3-1))
beta_D
```

```{r}
inv(t(D)%*%D)%*%t(D) %>% fractions
```

Okay, so what is this 1/4 doing here? There are only three data values in each group. You need to rewrite

1/4 = 1/3 - 1/12

Then the -1/12 represents subtracting the overall mean of all 12 observations. So the estimates involve an average of a group minus the overall average. In other words,

$\hat{\beta}_0=\bar{Y}_.$

$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_.$

$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_.$

$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_.$

where $\bar{Y}_.$ is the overall mean.

Again, this seems counter-intuitive, but remember that the interpretation of the regression coefficient in a multivariate setting is the effect of a one unit change in a variable **holding all the other variables constant**.

### Recode

Suppose you want to recombine the columns of your X matrix to transform it using some linear combinations. This is the same as post-multiplying by a matrix.

$B=AV$

or equivalently

$A=BV^{-1}$

So to convert from

```{r}
A
```

into

```{r}
B
```

you would multiply by this V

```{r, echo=FALSE}
V <- matrix(c(
  1,  0,  0,  0,
  1,  1,  0,  0,
  1,  0,  1,  0,
  1,  0,  0,  1), nrow=4, byrow=TRUE)
V
```

to get

```{r}
A%*%V
```

Note the V is not a symmetric matrix. Symmetry is not requirement here.

You can go in the other direction using

```{r}
U=inv(V)
U
```

which yields

```{r}
B%*%U
```

Now the formula for the estimates for a given parameterization is

$\hat{\beta}_B=(B'B)^{-1}B'Y$

Replace B with AV to get

$\hat{\beta}_B=(V'A'AV)^{-1}V'A'Y$

which simplifies, after a bit of work, to

$\hat{\beta}_B=V^{-1}(A'A)^{-1}A'Y$

or

$\hat{\beta}_B=V^{-1}\hat{\beta}_A$

Let's check this out with the coefficients we calculated earlier.

```{r}
beta_A
```

```{r}
beta_B
```

```{r}
inv(V)%*%beta_A
```

You can also go in the reverse direction.

```{r}
V%*%beta_B
```

### Check the deviation coding

The same works with any coding. The V matrix that gets you from

```{r}
A
```

to

```{r}
D
```

is

```{r, echo=FALSE}
V <- matrix(c(
  1, -1, -1, -1,
  1,  1,  0,  0,
  1,  0,  1,  0,
  1,  0,  0,  1), nrow=4, byrow=TRUE)
V
```

Let's check this in one direction

```{r}
A%*%V
```

and the opposite direction

```{r}
U=inv(V)
U
```

```{r}
D%*%U
```

We can establish that

```{r}
beta_A
```

can be converted to

```{r}
beta_D
```

using

```{r}
inv(V)%*%beta_A
```

and the reverse direction works as well.

```{r}
V%*%beta_D
```

### Make your own contrast

Suppose you wanted to develop a parameterization that is meaningful for your particular setting and none of the ones discussed here or elsewhere seem to fit the bill. With a bit of work, you can calculate your own.

Suppose for example that you wanted

$\hat{\beta}_0=\bar{Y}_1$

$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_1$

$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_2$

$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_3$

Instead of comparing each mean to the first mean, you want to compare each mean to the previous mean.

The U matrix that converts

$\hat{\beta}_0=\bar{Y}_1$

$\hat{\beta}_1=\bar{Y}_2$

$\hat{\beta}_2=\bar{Y}_3$

$\hat{\beta}_3=\bar{Y}_4$

to the new parameterization is

```{r, echo=FALSE}
U <- matrix(c(
  1,  0,  0,  0,
 -1,  1,  0,  0,
  0, -1,  1,  0,
  0,  0, -1,  1), nrow=4, byrow=TRUE)
U
```

Now, you want to travel from the parameterization represented by

```{r}
A
```

to a new parameterization represented by E. This would need V, the inverse of U, which is

```{r}
V <- inv(U)
V
```

which provides a matrix

```{r}
E <- A%*%inv(U)
E
```

which gives the correct results.

```{r}
x0 <- E[ , 1]
x1 <- E[ , 2]
x2 <- E[ , 3]
x3 <- E[ , 4]
beta_E <- coef(lm(y~x0+x1+x2+x3-1))
beta_E
```
### Acknowledgements
